---
title: "Installing Voyager"
output: html_document
vignette: >
  %\VignetteIndexEntry{Install}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

To install the release version of Voyager from Bioconductor, make sure that your Bioconductor version is 3.16 or higher, which requires R 4.2. 

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(version = "3.16") # Or a higher version in the future
BiocManager::install("Voyager")
```

While Voyager itself does not require compilation, some of its dependencies such as [`sf`](https://r-spatial.github.io/sf/) do, and require some system dependencies to compile. See the [Installation page of `sf`](https://r-spatial.github.io/sf/) for instructions to install the system dependencies.

Development version of Voyager can be installed from Bioconductor:
```{r}
BiocManager::install("Voyager", version = "devel")
```

Or from GitHub:
```{r}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("pachterlab/Voyager", ref = "devel")
```

