---
title: "Installing Voyager"
output: html_document
vignette: >
  %\VignetteIndexEntry{Install}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```
Users that are intersted in using the Python implementation of Voyager should reference the [VoyagerPy](https://pmelsted.github.io/voyagerpy/install/) website. 

To install the release version of Voyager from Bioconductor, make sure that your Bioconductor installation is version 3.16 or higher (this requires R version 4.2 or higher). 

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(version = "3.16") # Or a higher version in the future
BiocManager::install("Voyager")
```

While Voyager itself does not require compilation, some of its dependencies, such as [`sf`](https://r-spatial.github.io/sf/), must be compiled, and in turn rely on system dependencies. See the [Installation page of `sf`](https://r-spatial.github.io/sf/) for instructions on how to install the system dependencies.

The `main` branch on GitHub is the release version. The development version of Voyager can be installed from Bioconductor
```{r}
BiocManager::install("Voyager", version = "devel")
```

or directly from GitHub, from the `devel` branch:
```{r}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("pachterlab/voyager", ref = "devel")
```

