---
title: "Visium preprocessing with cellatlas"
author: "Kayla Jackson and A. Sina Booeshaghi"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: true
  toc_float:
    collapsed: false
  toc_depth: 3
jupyter:
  kernelspec:
    display_name: R
    language: R
    name: ir
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{visium_ca}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r, include = FALSE, eval = FALSE}
# Install Google Colab dependencies
install.packages(c("stringr", "rjson"))
```

```{r}
library(stringr)
library(rjson)
```

# Building Count Matrices with `cellatlas`
A major challenge in uniformly preprocessing large amounts of single-cell genomics data from a variety of different assays is identifying and handling sequenced elements in a coherent and consistent fashion. Cell barcodes in reads from RNAseq data from 10x Multiome, for example, must be extracted and error corrected in the manner as cell barcodes in reads from ATACseq data from 10x Multiome so that barcode-barcode registration can occur. Uniform processing in this way minimzes computational variability and enables cross-assay comparisons.

In this notebook we demonstrate how single-cell genomics data can be preprocessed to generate a cell by feature count matrix. This requires:

1. FASTQ files
2. `seqspec` specification for the FASTQ files
3. Genome Sequence FASTA
4. Genome Annotation GTF
5. (optional) Feature barcode list

# Install Packages
The vignette makes use of two non-standard command line tools, [`jq`](https://jqlang.github.io/jq/) and [`tree`](https://mama.indstate.edu/users/ice/tree/). The code cell below installs these tools on a Linux operating system and should be updated for Mac and Windows users.
```{r}
# Install `jq`, a command-line tool for extracting key value pairs from JSON files 
system("wget --quiet --show-progress https://github.com/stedolan/jq/releases/download/jq-1.6/jq-linux64")
system("chmod +x jq-linux64 && mv jq-linux64 /usr/local/bin/jq")
```
We will continue with other dependencies that can be installed on any operating system. 
```{r}
# Clone the cellatlas repo and install the package
system("git clone https://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@github.com/cellatlas/cellatlas.git")
system("cd cellatlas && pip install .")

# Install dependencies
system("yes | pip uninstall --quiet seqspec")
system("pip install --quiet git+https://github.com/IGVF/seqspec.git")
system("pip install --quiet gget kb-python")
```

# Preprocessing for Visium

## Examine the spec
**Note:** We move the relevant data to the working directory and `gunzip` the barcode onlist.
```{r}
system("mv cellatlas/examples/rna-visium-spatial/* .")
```
We first use `seqspec print` to check that the read structure matches what we expect. This command prints out an ordered tree representation of the sequenced elements contained in the FASTQ files. Note that the names of the nodes in the `seqspec` must match the names of the FASTQ files. Note that on Google Colab, go to Runtime -> View runtime logs to see the output from `system`.
```{r}
system("seqspec print spec.yaml")
```
## Fetch the references
This step is only necessary if the modality that we are processing uses a transcriptome reference-based alignment. 
```{r}
system("gget ref -o ref.json -w dna,gtf mus_musculus")
```
## Build the pipeline
```{r}
FA <- system2("jq",
  args = c("-r", "'.mus_musculus.genome_dna.ftp'", "ref.json"),
  stdout = TRUE)

GTF <- system2("jq",
  args = c("-r", "'.mus_musculus.annotation_gtf.ftp'", "ref.json"),
  stdout = TRUE)
```
We now supply all of the relevant objects to `cellatlas build` to produce the appropriate commands to be run to build the pipeline. This includes a reference building step and a read counting and quantification step both of which are performed with `kallisto` and `bustools` as part of the `kb-python` package.
```{r}
args <- c(
  "-o out", 
  "-s spec.yaml",
  "-m rna",  
  "-fa", FA,
  "-g", GTF,
  "fastqs/R1.fastq.gz fastqs/R2.fastq.gz")

system2(command = "cellatlas", args = c("build", args))
```
## Run the pipeline
We can extract and view the commands in the pipeline using `jq`. 
```{r}
cmds <- system2("jq", "-r '.commands[] | values[]' out/cellatlas_info.json", stdout=TRUE)
cmds <- str_subset(cmds, "[\\[\\]]", negate=TRUE)
cmds <- str_extract(cmds, "kb.*(txt|gz)")

cmds
```
Now we can run the commands from `out/cellatlas_info.json` on the command line. 
```{r}
lapply(cmds, function(cmd) system(cmd))
```
## Inspect the output 
We inspect the `out/run_info.json` and `out/kb_info.json` as a simple QC on the pipeline.
```{r}
list.files("out")
```

```{r}
rjson::fromJSON(file = "out/run_info.json")
```
